---
title: |
  <center> Incomplete Data Analysis </center>
  <center> EM algorithm for the two-component Gaussian mixture </center>
author: "V. Inácio de Carvalho & M. de Carvalho"
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document: default
  pdf_document: default
---
Here we implement the EM algorithm for the two-component Gaussian mixture model. Remember that for the E-step we have
\begin{align*}
E[Z_i\mid y,\theta^{(t)}]&=E[Z_i\mid y_i,\theta^{(t)}]\\
&=1\times\Pr(Z_i=1\mid y_i,\theta^{(t)})+0\times\Pr(Z_i=0\mid y_i,\theta^{(t)})\\
&=\frac{p^{(t)}\phi\left(y_i;\mu_1^{(t)},(\sigma_1^{(t)})^{2}\right)}{p^{(t)}\phi\left(y_i;\mu_1^{(t)},(\sigma_1^{(t)})^{2}\right)+(1-p^{(t)})\phi\left(y_i;\mu_2^{(t)},(\sigma_2^{(t)})^{2}\right)}\\
&=\widetilde{p}_{i}^{(t)},
\end{align*}
and thus
\begin{equation*}
Q(\theta\mid\theta^{(t)})=\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}\left\{\log p+\log \phi(y_i;\mu_1,\sigma_1^2)\right\}+\sum_{i=1}^{n}\left(1-\widetilde{p}_{i}^{(t)}\right)\left\{\log (1-p)+\log \phi(y_i;\mu_2,\sigma_2^2)\right\}.
\end{equation*}
For the M-step we have
\begin{align*}
p^{(t+1)}&=\frac{\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}}{n},\\
\mu_{1}^{(t+1)}&=\frac{\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}y_{i}}{\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}},\\
(\sigma_{1}^{(t+1)})^{2}&=\frac{\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}(y_{i}-\mu_{1}^{(t+1)})^{2}}{\sum_{i=1}^{n}\widetilde{p}_{i}^{(t)}},\\
\mu_{2}^{(t+1)}&=\frac{\sum_{i=1}^{n}(1-\widetilde{p}_{i}^{(t)})y_{i}}{\sum_{i=1}^{n}(1-\widetilde{p}_{i}^{(t)})},\\
(\sigma_{2}^{(t+1)})^{2}&=\frac{\sum_{i=1}^{n}(1-\widetilde{p}_{i}^{(t)})(y_{i}-\mu_{2}^{(t+1)})^{2}}{\sum_{i=1}^{n}(1-\widetilde{p}_{i}^{(t)})}.
\end{align*}
```{r, include = TRUE, message = FALSE}
em.mixture.two.normal <- function(y, theta0, eps){
n <- length(y)

theta <- theta0

p <- theta[1]
mu1 <- theta[2]; sigma1 <- theta[3]
mu2 <- theta[4]; sigma2 <- theta[5]

diff <- 1
while(diff > eps){

theta.old <- theta

#E-step
ptilde1 <- p*dnorm(y, mean = mu1, sd = sigma1)
ptilde2 <- (1 - p)*dnorm(y, mean = mu2, sd = sigma2)
ptilde <- ptilde1/(ptilde1 + ptilde2)

#M-step
p <- mean(ptilde)

mu1 <- sum(y*ptilde)/sum(ptilde)
sigma1 <- sqrt(sum(((y - mu1)^2)*ptilde)/sum(ptilde))

mu2 <- sum(y*(1 - ptilde))/sum(1 - ptilde)
sigma2 <- sqrt(sum(((y - mu2)^2)*(1 - ptilde))/sum(1 - ptilde))

theta <- c(p, mu1, sigma1, mu2, sigma2)
diff <- sum(abs(theta - theta.old))
}
return(theta)
}
```

We will now load the old faithful data set.
```{r, include = TRUE, message = FALSE}
data(faithful)
attach(faithful)
hist(waiting, main = "Time between Old Faithful eruptions",
     xlab = "Minutes",
     ylab="", cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.4)
```

Lastly, we apply the developed function to the old faithful data. For the starting values we have considered
\begin{equation*}
p^{(0)} = 0.4, \quad \mu_1^{(0)} = 40, \quad \sigma_1^{(0)} = 4, \quad \mu_2^{(0)} = 90, \quad \sigma_2^{(0)} = 4,
\end{equation*}
while for the stopping criterion we have used
\begin{equation*}
\lvert p^{(t+1)}-p^{(t)}\rvert + \lvert\mu_1^{(t+1)}-\mu_1^{(t)}\rvert + \lvert\sigma_1^{(t+1)}-\sigma_1^{(t)}\rvert + \lvert\mu_2^{(t+1)}-\mu_2^{(t)}\rvert + \lvert\sigma_2^{(t+1)}-\sigma_2^{(t)}\rvert < \varepsilon,
\end{equation*}
with $\varepsilon = 0.00001$
```{r, include = TRUE, message = FALSE}
res <- em.mixture.two.normal(y = waiting, c(0.4, 45, 4, 90, 4), 0.00001)
p <- res[1]
mu1 <- res[2]
sigma1 <- res[3]
mu2 <- res[4]
sigma2 <- res[5]

p; mu1; mu2; sigma1; sigma2

hist(waiting, main = "Time between Old Faithful eruptions",
     xlab = "Minutes", 
     ylab = "Density", 
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.4,
     freq = F, ylim = c(0,0.05))
curve(p*dnorm(x, mu1, sigma1) + (1 - p)*dnorm(x, mu2, sigma2),
      add = TRUE, lwd = 2, col = "blue2")
```

The package \texttt{mixtools} is capable of handling mixtures of Gaussians through the function \texttt{normalmixEM}. The code follows below.
```{r, include = TRUE, message = FALSE}
require(mixtools)
res1 <- normalmixEM(waiting, lambda = 0.4, mu = c(45, 90), sigma = c(4, 4))
p <- res1$lambda[1]
mu1 <- res1$mu[1]
mu2 <- res1$mu[2]
sigma1 <- res1$sigma[1]
sigma2 <- res1$sigma[2]

hist(waiting, main = "Time between Old Faithful eruptions",
     xlab = "Minutes",
     ylab = "Density",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.4,
     freq = F, ylim = c(0,0.05))
curve(p*dnorm(x, mu1, sigma1)+(1 - p)*dnorm(x, mu2, sigma2),
      add = TRUE, lwd = 2, col = "blue2")
```
