---
title: |
  <center> EM algorithm: (toy) examples </center>
author: "V. Inácio de Carvalho & M. de Carvalho"
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document: default
  pdf_document: default
---

In this supplementary file we implement the EM algorithm for the examples seen in the lecture. 

# Toy example (2 exponential observations)

Remember that we have the following updating equation
\begin{equation*}
\theta^{(t+1)}=\frac{2\theta^{(t)}}{5\theta^{(t)}+1}.
\end{equation*}

Note that at convergence, $\theta^{(t)}\rightarrow \theta^{(t+1)}\rightarrow \widehat{\theta}$, and so a fixed point of these iterations is $\widehat{\theta}=1/5$. Nevertheless, let us code it. The function takes as input a starting value for $\theta$, say $\theta^{(0)}$, and the value $\epsilon$ used for the stopping criterion: $|\theta^{(t)}-\theta^{(t-1)}|<\epsilon$. The variable \texttt{diff} in the code below is just to control whether the convergence criterion is met or not. Although we have started setting \texttt{diff=1}, any value greater than $\epsilon$ would work. The variable \texttt{theta.old} stores the value from the previous iteration, so that we can compute $|\theta^{(t+1)}-\theta^{(t)}|$.

```{r, include = TRUE, message = FALSE}
toyex <- function(theta0, eps){

diff <- 1
theta <- theta0
while(diff > eps){
  
theta.old <- theta  
theta <- 2*theta/(5*theta+1)
diff <- abs(theta - theta.old)
}
return(theta)
}

toyex(10, 0.00001)
```

# Genetic linkage model
Remember that we have for the E-step
\begin{align*}
Q(\theta\mid\theta^{(t)})&=(y_1-z^{(t)}+y_4)\log \theta+(y_2+y_3)\log(1-\theta),\\
z^{(t)}&=y_1\times\frac{1/2}{1/2+\theta^{(t)}/4},
\end{align*}
while for the M-step we have
\begin{equation*}
\theta^{(t+1)}=\frac{y_1-z^{(t)}+y_4}{n-z^{(t)}}, \quad n=y_1+y_2+y_3+y_4.
\end{equation*}

```{r, include = TRUE, message = FALSE}
multi <- function(y, theta0, eps){
n <- sum(y); diff <- 1
theta <- theta0

while(diff>eps){
theta.old <- theta

#E step
zt <- y[1]*0.5/(0.5 + 0.25*theta)

#M step
theta <- (y[1] + y[4] - zt)/(n - zt)

diff <- abs(theta - theta.old)
}
return(theta)
}

y <- c(125, 18, 20, 34)
multi(y = y, 0.5, 0.00001)
```

# Incomplete univariate (normal) data

Remember that we have the following updating equation
\begin{equation*}
\mu^{(t+1)}=\frac{\sum_{i=1}^{m}y_i+(n-m)\mu^{(t)}}{n}.
\end{equation*}

Again, at convergence, $\mu^{(t)}\rightarrow\mu^{(t+1)}\rightarrow \widehat{\mu}$ and so a fixed point of these iterations is $\widehat{\mu}=\frac{1}{m}\sum_{i=1}^{m}y_i$, exactly what we would have obtained by maximising the log likelihood of the observed data.

For this example I have simulated $n=200$ observations and the true value of $\mu$ is $3$. The missing data mechanism is MCAR and I have simply sampled 20 individuals to exclude from the analysis. So, in the notation of our example we have $n=200$ and $m=180$.

```{r, include = TRUE, message = FALSE}
n <- 200
mu <- 3
set.seed(1)
y <- rnorm(n, mu, 1)
ind <- sample(x = 1:n, size = 20, replace = FALSE)

#observed data
y_obs <- y[-ind]

toyex <- function(mu0, eps, y, n){

diff <- 1
mu <- mu0
m <- length(y)
  
while(diff > eps){
    mu.old <- mu 
    mu <-  (sum(y) + (n-m)*mu)/n
    diff <-  abs(mu - mu.old)
  }
  return(mu)
}

toyex(15, 0.00001, y = y_obs, n = n)

sum(y_obs)/(n-length(ind))
```
