---
title: |
  <center> Incomplete Data Analysis </center>
  <center> Ignorability---simulation study </center>
author: "V. Inácio de Carvalho & M. de Carvalho"
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document: default
  pdf_document: default
---

In this supplementary file I show how to reproduce the results from the simulation study presented in the slides.  We are assuming that $Y_i\overset{\text{iid}}\sim \text{Bernoulli}(\theta)$ and $R_i\mid Y_i\overset{\text{iid}}\sim\text{Bernoulli}(\theta)$. I will simulate $nsim=1000$ datasets of sample size $n=100$ and consider $\theta=0.3$. I will store the generated data and corresponding missing data indicators in a $n\times nsim$ matrix. The maximum likelihood estimates (from the 1000 simulated datasets) based on both the full and observed data likelihood will be stored in two separate vectors.

```{r, include = TRUE, message = FALSE}
nsim <- 1000; n <- 100; theta <- 0.3
y <- r <- matrix(0, nrow = n, ncol = nsim)
mle_full <- mle_observed <- numeric(nsim)

set.seed(1)
for(l in 1:nsim){
y[, l] <- rbinom(n, 1, theta)
r[, l] <- rbinom(n, 1, theta) 

m <- length(which(r[,l] == 1))

mle_full[l] <- (m + sum(y[r[, l] == 1, l]))/(m + n)
mle_observed[l] <- sum(y[r[, l] == 1, l])/m
}

hist(mle_full, ylim = c(0, 250), xlim = c(0, 0.6),
     xlab = expression(theta[mle]), main = "Full likelihood")
abline(v = theta, col = "red", lwd = 3)

hist(mle_observed, ylim = c(0, 250), xlim = c(0, 0.6),
     xlab = expression(theta[mle]), main = "Observed data likelihood")
abline(v = theta, col="red", lwd = 3)
```

We will now violate the MAR assumption. We repeat a similar exercise but now, instead of violating the non-distinctness of parameters assumption, we violate the MAR assumption. In particular, we assume
\begin{equation*}
Y_i\overset{\text{iid}}\sim\text{Bernoulli}(\theta),\quad \text{and}\quad \Pr(R_i=1\mid Y_i)=\frac{e^{Y_i}}{1+e^{Y_i}}.
\end{equation*}
The code follows below.

```{r, include = TRUE, message = FALSE}
r <- matrix(0, nrow = n, ncol = nsim)
mle_observed_mnar <- numeric(nsim)

set.seed(1)
for(l in 1:nsim){
r[, l] <- rbinom(n, 1, exp(y[, l])/(1+exp(y[, l]))) 

m <- length(which(r[,l] == 1))

mle_observed_mnar[l] <- sum(y[r[, l] == 1, l])/m
}

hist(mle_observed_mnar, ylim = c(0, 400), xlim = c(0, 0.6),
     xlab = expression(theta[mle]), main = "MNAR data")
abline(v = theta, col = "red", lwd = 3)
```

