---
title: "test regression house prices"
output: html_document
date: "2023-06-30"
---

---
title: "House Prices India Notebook"
output: html_notebook
---
#LIBRARIES
```{r}

library("readr")
library("scales")
library("dplyr")
library("corrplot")
library("vioplot")
library("tidyr")
library("factoextra")
library("GGally")
library("dplyr")
#library("ggplot")
library("qwraps2")
library("lubridate")
library("dbscan")
library("ggplot2")
library("gdtools")
library("Cairo")
library("stringr")
library("data.table")
library("modelr")
library("DataExplorer")
library("corrplot")
library("plotly")
library("nortest")
library("ggrepel")
library("raster")
library("sp")
#library("rgeos")
#library("spTransform")
library("dplyr")
library("caret")
library("rpart")
#library("ggmap")
library("tidyr")
library("vioplot")
```

Column	Description
POSTED_BY	Category marking who has listed the property
UNDER_CONSTRUCTION	Under Construction or Not
RERA	Rera approved or Not. Note: RERA stands for Real Estate Regulatory Authority. It is a regulatory body established in many countries to oversee and regulate the real estate sector
BHK_NO	Number of Rooms
BHK_OR_RK	Type of property. Note: BHK: BHK stands for Bedroom, Hall, and Kitchen; K stands for Room Kitchen. It is used to describe smaller residential unit
SQUARE_FT	Total area of the house in square feet
READY_TO_MOVE	Category marking Ready to move or Not
RESALE	Category marking Resale or not
ADDRESS	Address of the property
LONGITUDE	Longitude of the property            
LATITUDE	Latitude of the property                 
price_LACs Target price of the property         


#READ FILE / DATAFRAME
```{r}
setwd("C:/Users/manue/Downloads/House Price Prediction Challenge/dataset")
house_prices = read_csv('./train.csv')
DataExplorer::create_report(house_prices)

dim(house_prices)    #12 features
str(house_prices)
summary(house_prices)
```

#Missing values?
```{r}
sum(is.na(house_prices))

```
```{r}
# Define the exchange rate (example value)
exchange_rate <- 1116.32#1113.69#0.011

# Create a new column with the target price in euros
#house_prices$target_price_eur <- house_prices$TARGET(PRICE_IN_LACS) * exchange_rate
house_prices$target_price_eur <- house_prices$`TARGET(PRICE_IN_LACS)` * exchange_rate
```

#LATITUDE/LONGITUDE - Negative values outside range of India territory (?!)
```{r}

# Create price categories based on ranges
house_prices$price_category <- cut(house_prices$TARGET.PRICE_IN_LACS,
                                   breaks = c(0, 5000, 10000, 20000, Inf),include.lowest = T, right = TRUE, dig.lab = 3,
    ordered_result = FALSE,
                                   labels = c(">5000", ">10000", ">20000",">30000"))

ggplot(house_prices, aes(x = LONGITUDE, y = LATITUDE, color = price_category)) +
  geom_point(alpha = 0.5) +
  theme_minimal()

house_prices$LATITUDE <- NULL
house_prices$LONGITUDE <- NULL
house_prices$price_category <- NULL
str(house_prices)
```
# BINARY FEATURES
```{r}

proportions <- prop.table(table(house_prices$UNDER_CONSTRUCTION))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", scales::percent(proportions), ")")
pie(proportions, labels = labels, main = "Under Construction or Not")

proportions <- prop.table(table(house_prices$RERA))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Rera approved or No")

proportions <- prop.table(table(house_prices$READY_TO_MOVE))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Ready to move")

proportions <- prop.table(table(house_prices$RESALE))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Resale property")

house_prices$RESALE <- as.numeric(house_prices$RESALE)
house_prices$READY_TO_MOVE <- as.numeric(house_prices$READY_TO_MOVE)
house_prices$RERA <- as.numeric(house_prices$RERA)
house_prices$UNDER_CONSTRUCTION <- as.numeric(house_prices$UNDER_CONSTRUCTION)
```

#NUMBER OF ROOMS - CHANGE OF VARIABLE TO NUMERIC
```{r}
house_prices$BHK_NO. <- as.numeric(house_prices$BHK_NO.)
```


#Advertised by "Builder" = 1, "Owner" = 2 "Dealer" = 3 -change of variables            (nota: não correr 2 vezes, gera uma coluna de NA's)
```{r}
summary(house_prices$POSTED_BY)
summary_posted1 <- house_prices %>%
  count(POSTED_BY, name = "count_postedby") %>%
  arrange(desc(count_postedby))
summary_posted1

house_prices$POSTED_BY <- trimws(house_prices$POSTED_BY)
house_prices <- house_prices %>%
  mutate(POSTED_BY  = if_else(POSTED_BY  == "Builder", 1, if_else(POSTED_BY == "Owner", 3, 2)))



summary_posted2<- house_prices%>%
  count(POSTED_BY, name = "count_posts") %>%
  arrange(desc(count_posts))
summary_posted2

house_prices$POSTED_BY <- as.numeric(house_prices$POSTED_BY)
class(house_prices$POSTED_BY )
str(house_prices$POSTED_BY)
summary(house_prices$POSTED_BY)


```

#ADDRESS
```{r}
table(house_prices$ADDRESS)
house_prices$ADDRESS<- gsub(",","",house_prices$ADDRESS)
house_prices$ADDRESS<- gsub(" ","",house_prices$ADDRESS)

summary_street <- house_prices %>%
  count(ADDRESS, name = "count_adress") %>%
  arrange(desc(count_adress))
summary_street
str(summary_street)

addresses <- unique(house_prices$ADDRESS)
numeric_addresses <- numeric(length(house_prices$ADDRESS))

for (i in 1:length(addresses)) {
  numeric_addresses[house_prices$ADDRESS == addresses[i]] <- i
}
house_prices$ADDRESS <- numeric_addresses
#View(house_prices)
```

#Type of house - change to numeric variables - BHK = 1   RK = 0
```{r}
table(house_prices$BHK_OR_RK)  # BHK    RK  29427    24 

house_prices$BHK_OR_RK <- trimws(house_prices$BHK_OR_RK)
house_prices <- house_prices %>%
  mutate(BHK_OR_RK  = if_else(BHK_OR_RK  == "BHK", 1, 0))
colnames(house_prices)
```

# Create a new column with the target price in euros
```{r}

exchange_rate <- 1 /0.011  #0.011
exchange_rate

house_prices$target_price_eur <- house_prices$`TARGET.PRICE_IN_LACS.`* exchange_rate

house_prices$TARGET.PRICE_IN_LACS.<- NULL
View(house_prices)
```
# TRANSFORM SQUARE Feet TO SQUARe METER (1 -> 0.09290303997)
```{r}

house_prices$SQUARE_MT <- house_prices$SQUARE_FT * 0.09290303997
house_prices$SQUARE_FT <- NULL
  
```

```{r}
house_prices_regression <- house_prices[, !(names(house_prices) %in% c("POSTED_BY", "BHK_OR_RK", "ADDRESS", "TARGET(PRICE_IN_LACS)"))]
# Convert the entire data frame to numeric
house_prices_regression <- as.data.frame(lapply(house_prices_regression, as.numeric))

# Check the data types
str(house_prices_regression)





```



```{r}
# Display column names and data types
str(house_prices_regression)

# Calculate non-null counts
non_null_counts <- sapply(house_prices_regression, function(x) sum(!is.na(x)))
print(non_null_counts)

# Calculate summary statistics
summary_stats <- sapply(house_prices_regression, summary)
print(summary_stats)

```


```{r}
# Assign the target_price_eur column to y
y <- house_prices_regression$target_price_eur

# Create X by dropping the target_price_eur column
X <- house_prices_regression[, !(names(house_prices_regression) %in% c("target_price_eur"))]

# Display the head of X
head(X)

```

```{r}
# Load the caret package
library(caret)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression), 0.7 * nrow(house_prices_regression))
train <- house_prices_regression[train_indices, ]
test <- house_prices_regression[-train_indices, ]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(linear_model, newdata = test)
linear_mse <- mean((linear_predictions - test$target_price_eur)^2)

# Print the mean squared error (MSE)
print(linear_mse)

```
Removing LAT and LONG

```{r}
house_prices_regression2 <- house_prices[, !(names(house_prices) %in% c("POSTED_BY", "BHK_OR_RK", "ADDRESS", "TARGET(PRICE_IN_LACS)", "LONGITUDE", "LATITUDE"))]
# Convert the entire data frame to numeric
house_prices_regression2 <- as.data.frame(lapply(house_prices_regression2, as.numeric))

# Check the data types
str(house_prices_regression2)
```
```{r}
# Display column names and data types
str(house_prices_regression2)

# Calculate non-null counts
non_null_counts <- sapply(house_prices_regression2, function(x) sum(!is.na(x)))
print(non_null_counts)

# Calculate summary statistics
summary_stats <- sapply(house_prices_regression2, summary)
print(summary_stats)


```
```{r}
# Assign the target_price_eur column to y
y <- house_prices_regression2$target_price_eur

# Create X by dropping the target_price_eur column
X <- house_prices_regression2[, !(names(house_prices_regression2) %in% c("target_price_eur"))]

# Display the head of X
head(X)
```
```{r}
# Load the caret package
library(caret)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression2), 0.7 * nrow(house_prices_regression2))
train <- house_prices_regression2[train_indices, ]
test <- house_prices_regression2[-train_indices, ]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(linear_model, newdata = test)
linear_mse <- mean((linear_predictions - test$target_price_eur)^2)

# Print the mean squared error (MSE)
print(linear_mse)
```
```{r}
colnames(house_prices_regression2)

feats <- c("UNDER_CONSTRUCTION","RERA","BHK_NO.","READY_TO_MOVE","RESALE","SQUARE_MT")
target <- "target_price_eur"
```
```{r}
# Split the data into features (X) and target variable (y)
X <- house_prices_regression2[, feats]  # Exclude the last column if it represents the target variable
y <- house_prices_regression2[, target]   # Select the last column as the target variable

# Split the data into training and testing sets
set.seed(42)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression2), 0.7 * nrow(house_prices_regression2))
train <- house_prices_regression2[train_indices, c(feats,target) ]
test <- house_prices_regression2[-train_indices,  c(feats,target)]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Train a regression tree model
#tree_model <- rpart(popularity ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(linear_model, newdata = test)
linear_mse <- mean((linear_predictions - unlist(test[,target]))^2)
print(linear_mse)
```
```{r}
colnames(house_prices_regression)

feats <- c("UNDER_CONSTRUCTION","RERA","BHK_NO.","READY_TO_MOVE","RESALE","LONGITUDE","LATITUDE","SQUARE_MT")
target <- "target_price_eur"
```
```{r}
# Split the data into features (X) and target variable (y)
X <- house_prices_regression[, feats]  # Exclude the last column if it represents the target variable
y <- house_prices_regression[, target]   # Select the last column as the target variable

# Split the data into training and testing sets
set.seed(42)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression), 0.7 * nrow(house_prices_regression))
train <- house_prices_regression[train_indices, c(feats,target) ]
test <- house_prices_regression[-train_indices,  c(feats,target)]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Train a regression tree model
#tree_model <- rpart(popularity ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(linear_model, newdata = test)
linear_mse <- mean((linear_predictions - unlist(test[,target]))^2)
print(linear_mse)
```
```{r}
# Split the data into features (X) and target variable (y)
X <- house_prices_regression[, feats]  # Exclude the last column if it represents the target variable
y <- house_prices_regression[, target]   # Select the last column as the target variable

# Split the data into training and testing sets
set.seed(42)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression), 0.7 * nrow(house_prices_regression))
train <- house_prices_regression[train_indices, c(feats,target) ]
test <- house_prices_regression[-train_indices,  c(feats,target)]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Train a regression tree model
tree_model <- rpart(target_price_eur ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(linear_model, newdata = test)
linear_mse <- mean((linear_predictions - unlist(test[,target]))^2)

tree_predictions <- predict(tree_model, newdata = test)
tree_mse <- mean((tree_predictions - unlist(test[,target]))^2)
#print(linear_mse)

summary(linear_model) # ver as feature_importance

baseline <-mean(unlist(y))# ver a baseline para o erro
paste("Media Baseline: ", baseline)

baseline_mse <- mean((baseline - unlist(test[,target]))^2)
paste("MSE Baseline", baseline_mse)
paste("SQRT: MSE Baseline", sqrt(baseline_mse))

paste("__________\n\n\ MSE Linear: ", linear_mse)
paste("SQRT: MSE Baseline:", sqrt(linear_mse))

paste("__________\n\n\ MSE Linear: ", tree_mse)
paste("SQRT: MSE Baseline:", sqrt(tree_mse))


```

```{r}
library(knitr)

# Create a data frame with the data
data <- data.frame(
  Metric = c("MSE Baseline", "SQRT: MSE Baseline", "MSE Linear", "SQRT: MSE Linear", "MSE Tree", "SQRT: MSE Tree"),
  Value = c(baseline_mse, sqrt(baseline_mse), linear_mse, sqrt(linear_mse), tree_mse, sqrt(tree_mse))
)

# Print the table
paste(kable(data))

```
```{r}
library(knitr)

# Create a data frame with the data
data <- data.frame(
  Metric = c("MSE Baseline", "SQRT: MSE Baseline", "MSE Linear", "SQRT: MSE Linear", "MSE Tree", "SQRT: MSE Tree"),
  Value = c(baseline_mse, sqrt(baseline_mse), linear_mse, sqrt(linear_mse), tree_mse, sqrt(tree_mse))
)

# Print the table
paste(kable(data))
```



```{r}
# Split the data into features (X) and target variable (y)
X <- house_prices_regression[, feats]  # Exclude the last column if it represents the target variable
y <- house_prices_regression[, target]   # Select the last column as the target variable

# Split the data into training and testing sets
set.seed(42)  # For reproducibility
train_indices <- sample(1:nrow(house_prices_regression), 0.7 * nrow(house_prices_regression))
train <- house_prices_regression[train_indices, c(feats,target) ]
test <- house_prices_regression[-train_indices,  c(feats,target)]

# Train a linear regression model
linear_model <- lm(target_price_eur ~ ., data = train)

# Train a regression tree model
#tree_model <- rpart(target_price_eur ~ ., data = train)

# Evaluate the models using split validation
linear_predictions <- predict(tree_model, newdata = test)
linear_mse <- mean((linear_predictions - unlist(test[,target]))^2)
print(linear_mse)

summary(linear_model) # ver as feature_importance
baseline <-mean(unlist(y))# ver a baseline para o erro
baseline_mse <- mean((baseline - unlist(test[,target]))^2)
```
```{r}
baseline
```
```{r}
baseline_mse
```

