---
title: 
author: "V. Inácio & M. de Carvalho"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

Here I reproduce the plots we have in the slides. They are easily obtained using the package \texttt{survival}. The crux is to create an object of class \texttt{Surv} containing the event times and whether or not they are censored.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
require(survival)
times <- c(1, 3, 3, 6, 8, 9, 10)
status <- c(1, 1, 1, 0, 0, 1, 0)
S <- Surv(times, status)
S
class(S)
```

Note that when we print our object \texttt{S}, the censored times appear with a \texttt{+} symbol. We can then use the \texttt{survfit} function in order to obtain the Kaplan--Meier estimate of the survival function. Note that \texttt{conf.type = "plain"} implements the approach of slide 17 (based on Greenwood formula).
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), figure.width = 5, figure.height = 5}
fit <- survfit(S ~ 1, conf.type = "plain")
fit
summary(fit)
plot(fit,
     xlab = "Time (in months)", 
     ylab = "Kaplan-Meier estimate of the survival curve")
```

From the output we can see that the estimated median survival time is 9 (months). Other type of (pointwise) confidence intervals for the survival function can be obtained by changing the argument \texttt{conf.type}. A popular approach is to set it to \texttt{log-log}. 
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), figure.width = 5, figure.height = 5}
fit_alt <- survfit(S ~ 1, conf.type = "log-log")
summary(fit_alt)
plot(fit_alt,
     xlab = "Time (in months)", 
     ylab = "Kaplan-Meier estimate of the survival curve")
```

With regard to the Nelson--Aalen estimate, we proceed exactly in the same way as for the Kaplan--Meier estimate, we just need to specify the following arguments \texttt{stype = 2} and \texttt{ctype = 1}. More information can be find here:
\begin{center}
\texttt{https://stat.ethz.ch/R-manual/R-devel/library/survival/html/survfit.formula.html}
\end{center}

```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), figure.width = 5, figure.height = 5}
fit_NA <- survfit(S ~ 1, stype = 2, ctype = 1)
summary(fit_NA)
plot(fit_NA, conf.int = FALSE,
     xlab = "Time (in months)", 
     ylab = "Nelson-Aalen estimate of the survival curve")
```

Let us visualise both the Kaplan--Meier and the Nelson--Aalen estimates in the same plot.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), figure.width = 5, figure.height = 5}
plot(fit_NA, conf.int = FALSE,
     xlab = "Time (in months)", 
     ylab = "Estimate of the survival curve")
lines(fit, conf.int = FALSE, col = 2, lty = 2)
legend("topright", lty = c(1, 2), col = c(1, 2),
       c("Nelson-Aalen", "Kaplan-Meier"), bty = "n")
```

A recent package that produces very nice visualisations is \texttt{survminer}.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), figure.width = 5, figure.height = 5}
require(survminer)
toy_data <- data.frame("times" = times, "status" = status)
res_fit <- survfit(Surv(times, status) ~ 1, data = toy_data)
ggsurvplot(res_fit, 
           conf.int = 0.95,
           ggtheme = theme_minimal(), 
           data = toy_data,
           risk.table = TRUE)
```
