---
title: ""
author: "V. Inácio & M. de Carvalho"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---
Here we reproduce the results for the larynx data.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
require(KMsurv) #for the larynx data
require(survival)
data(larynx)
res_1 <- coxph(Surv(time, delta) ~ as.factor(stage) + age, data = larynx)
res_1
summary(res_1)
```

In order to apply the likelihood ratio test, let us fit a proportional hazards model that does not include the stage of the cancer as risk factor. Our null hypothesis is then: $H_0:\beta_1=\beta_2=\beta_3=0$ against the alternative that, at least, one of the coefficients is not zero.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
res_2 <- coxph(Surv(time, delta) ~ age, data = larynx)
res_2
anova(res_1,res_2)
```

We can also obtain this value manually as follows.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
#the first is the log likelihood of a model that contains none of the risk factors,
#so we need the second one
res_1$loglik
test_stat <- 2*(res_1$loglik[2] - res_2$loglik[2])
pchisq(test_stat, df = 4 - 1, lower.tail = FALSE)
```

If we want to compare non nested models, we can use the AIC or the BIC. 
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
AIC(res_1)
-2*res_1$loglik[2] + 2*4
BIC(res_1)
-2*res_1$loglik[2] + log(sum(larynx$delta))*4
```

Here we reproduce the results to obtain the adjusted survival curves (under a Cox model) presented in the last slide.
```{r, include = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60)}
require(survival)
require(KMsurv)
data(larynx)
res_1 <- coxph(Surv(time, delta) ~ as.factor(stage) + age, data = larynx)

#adjusted survival curve for the mean age
newdf <- data.frame("stage" = levels(as.factor(larynx$stage)), 
                    "age" = rep(mean(larynx$age), 4)) 
fit <- survfit(res_1, newdata = newdf)

plot(fit, conf.int = FALSE, col = c(1:4),
     ylab = "Estimated survival probability", 
     xlab = "Time (in years)", 
     main = "Age = 64.6 years (mean age)")
legend("topright", legend = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
       col = c(1:4), lty = c(1,1,1,1), bty = "n")
```
