---
title: "House Prices India Notebook"
output: html_notebook
---
#LIBRARIES
```{r}

library("readr")
library("scales")
library("dplyr")
library("corrplot")
library("vioplot")
library("tidyr")
library("factoextra")
library("GGally")
library("dplyr")
#library("ggplot")
library("qwraps2")
library("lubridate")
library("dbscan")
library("ggplot2")
library("gdtools")
library("Cairo")
library("stringr")
library("data.table")
library("modelr")
library("DataExplorer")
library("corrplot")
library("plotly")
library("nortest")
library("ggrepel")
library("raster")
library("sp")
library("rgeos")
#library("spTransform")
library("dplyr")
library("caret")
library("rpart")
library("ggmap")
library("tidyr")
library("vioplot")
```

Column	Description
POSTED_BY	Category marking who has listed the property
UNDER_CONSTRUCTION	Under Construction or Not
RERA	Rera approved or Not. Note: RERA stands for Real Estate Regulatory Authority. It is a regulatory body established in many countries to oversee and regulate the real estate sector
BHK_NO	Number of Rooms
BHK_OR_RK	Type of property. Note: BHK: BHK stands for Bedroom, Hall, and Kitchen; K stands for Room Kitchen. It is used to describe smaller residential unit
SQUARE_FT	Total area of the house in square feet
READY_TO_MOVE	Category marking Ready to move or Not
RESALE	Category marking Resale or not
ADDRESS	Address of the property
LONGITUDE	Longitude of the property            
LATITUDE	Latitude of the property                 
price_LACs Target price of the property         


#READ FILE / DATAFRAME
```{r}
house_prices=read.csv("C:/Users/mcosme.hbkworld/Downloads/housing prices/train.csv", stringsAsFactors = F)
#DataExplorer::create_report(house_prices)

dim(house_prices)    #12 features
str(house_prices)
summary(house_prices)
```

#Missing values?
```{r}
sum(is.na(house_prices))

```

# BINARY FEATURES
```{r}

proportions <- prop.table(table(house_prices$UNDER_CONSTRUCTION))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", scales::percent(proportions), ")")
pie(proportions, labels = labels, main = "Under Construction or Not")

proportions <- prop.table(table(house_prices$RERA))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Rera approved or No")

proportions <- prop.table(table(house_prices$READY_TO_MOVE))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Ready to move")

proportions <- prop.table(table(house_prices$RESALE))
proportions <- as.numeric(proportions)
labels <- paste0(c("0", "1"), " (", sprintf("%.1f%%", proportions*100), ")")
pie(proportions, labels = labels, main = "Resale property")

house_prices$RESALE <- as.numeric(house_prices$RESALE)
house_prices$READY_TO_MOVE <- as.numeric(house_prices$READY_TO_MOVE)
house_prices$RERA <- as.numeric(house_prices$RERA)
house_prices$UNDER_CONSTRUCTION <- as.numeric(house_prices$UNDER_CONSTRUCTION)
```

#NUMBER OF ROOMS - CHANGE OF VARIABLE TO NUMERIC
```{r}
house_prices$BHK_NO. <- as.numeric(house_prices$BHK_NO.)
head(house_prices)
```


#Advertised by "Builder" = 1, "Owner" = 2 "Dealer" = 3 -change of variables            (nota: não correr 2 vezes, gera uma coluna de NA's)
```{r}
summary(house_prices$POSTED_BY)
summary_posted1 <- house_prices %>%
  count(POSTED_BY, name = "count_postedby") %>%
  arrange(desc(count_postedby))
summary_posted1

house_prices$POSTED_BY <- trimws(house_prices$POSTED_BY)
house_prices <- house_prices %>%
  mutate(POSTED_BY  = if_else(POSTED_BY  == "Builder", 1, if_else(POSTED_BY == "Owner", 2, 3)))



summary_posted2<- house_prices%>%
  count(POSTED_BY, name = "count_posts") %>%
  arrange(desc(count_posts))
summary_posted2

house_prices$POSTED_BY <- as.numeric(house_prices$POSTED_BY)
class(house_prices$POSTED_BY )
str(house_prices$POSTED_BY)
summary(house_prices$POSTED_BY)


```

#ADDRESS
```{r}
table(house_prices$ADDRESS)
house_prices$ADDRESS<- gsub(",","",house_prices$ADDRESS)
house_prices$ADDRESS<- gsub(" ","",house_prices$ADDRESS)

summary_street <- house_prices %>%
  count(ADDRESS, name = "count_adress") %>%
  arrange(desc(count_adress))
summary_street
str(summary_street)

addresses <- unique(house_prices$ADDRESS)
numeric_addresses <- numeric(length(house_prices$ADDRESS))

for (i in 1:length(addresses)) {
  numeric_addresses[house_prices$ADDRESS == addresses[i]] <- i
}
house_prices$ADDRESS <- numeric_addresses
#View(house_prices)
```

#Type of house - change to numeric variables - BHK = 1   RK = 0
```{r}
table(house_prices$BHK_OR_RK)  # BHK    RK  29427    24 

house_prices$BHK_OR_RK <- trimws(house_prices$BHK_OR_RK)
house_prices <- house_prices %>%
  mutate(BHK_OR_RK  = if_else(BHK_OR_RK  == "BHK", 1, 0))
colnames(house_prices)
```

# Create a new column with the target price in euros 1 lac = 100000 rupees = 1116.48eur
```{r}

exchange_rate <- 1116.48  #0.011
exchange_rate

house_prices$target_price_eur <- house_prices$`TARGET.PRICE_IN_LACS.`* exchange_rate

house_prices$TARGET.PRICE_IN_LACS.<- NULL
summary(house_prices$target_price_eur)
View(house_prices)
```
#LATITUDE/LONGITUDE - Negative values outside range of India territory (?!)
```{r}

# Create price categories based on ranges
house_prices$price_category <- cut(house_prices$target_price_eur,
                                   breaks = c(0, 5000, 10000, 20000, Inf),include.lowest = T, right = TRUE, dig.lab = 3,
    ordered_result = FALSE,
                                   labels = c(">5000", ">10000", ">20000",">30000"))

ggplot(house_prices, aes(x = LONGITUDE, y = LATITUDE, color = price_category)) +
  geom_point(alpha = 0.5) +
  theme_minimal()

house_prices$LATITUDE <- NULL
house_prices$LONGITUDE <- NULL
house_prices$price_category <- NULL
str(house_prices)
```
#TRANSFORMAR LAT E LONG EM NORMAS
```{r}

```


# TRANSFORM SQUARE Feet TO SQUARe METER (1 -> 0.09290303997)
```{r}

house_prices$SQUARE_MT <- house_prices$SQUARE_FT * 0.09290303997
house_prices$SQUARE_FT <- NULL
  
```

```{r}
house_prices_regression <- house_prices
DataExplorer::create_report(house_prices_regression)
```

#TRANSFORMED DATA SET FOR scaling and outliers
```{r}
summary(house_prices)
str(house_prices)
View(house_prices)
```
```{r}
DataExplorer::create_report(house_prices)
```

#TEST DISTRIBUTION OF Target Price
```{r}

hp_sample <- house_prices %>% sample_n(4999)   
shapiro.test(hp_sample$target_price_eur)


house_prices %>%
  filter(target_price_eur >= 0L & target_price_eur <= 500000L) %>%
  ggplot() +
  aes(x = target_price_eur) +
  geom_histogram(bins = 45L, fill = "#112446") +
  theme_minimal()


```
#SCALING FEATURES - POSTED_BY 
```{r} 
#POSTED_BY

summary(house_prices$POSTED_BY)
table(house_prices$POSTED_BY)

scaled_house_prices <- house_prices %>%
  mutate(POSTED_BY  = if_else(POSTED_BY  == "1", 622/29451 , if_else(POSTED_BY == "2", 10538/29451, 18291/29451)))
head(scaled_house_prices)

head(scaled_house_prices, 20)
table(scaled_house_prices$POSTED_BY)

```

#ALL OTHER CONTINUOS FEATURES - as distribussion are Poisson we use min-max scaling
```{r}
View(scaled_house_prices)

scaled_house_prices$BHK_NO. = (scaled_house_prices$BHK_NO. - min(scaled_house_prices$BHK_NO.))/(max(scaled_house_prices$BHK_NO.) - min(scaled_house_prices$BHK_NO.))
scaled_house_prices$ADDRESS<- (scaled_house_prices$ADDRESS-min(scaled_house_prices$ADDRESS))/(max(scaled_house_prices$ADDRESS)-min(scaled_house_prices$ADDRESS))
scaled_house_prices$target_price_eur<- (scaled_house_prices$target_price_eur-min(scaled_house_prices$target_price_eur))/(max(scaled_house_prices$target_price_eur)-min(scaled_house_prices$target_price_eur))
scaled_house_prices$SQUARE_MT = (scaled_house_prices$SQUARE_MT - min(scaled_house_prices$SQUARE_MT))/(max(scaled_house_prices$SQUARE_MT) - min(scaled_house_prices$SQUARE_MT))

```

#CLUSTERING: without scaling
```{r}
resultado_kmeans <- kmeans(house_prices_regression, centers = 4)
house_prices_regression$cluster <- as.character(resultado_kmeans$cluster)


ggplot(scaled_house_prices) +
    aes(y = house_prices_regression$target_price_eur, x = house_prices_regression$SQUARE_MT  , colour = cluster) +
    geom_point(shape = "circle", size = 1.5) +
    scale_color_hue(direction = 1) +
    theme_minimal() +
    coord_cartesian(xlim = c(0, 25000), ylim = c(0, 15000000))

head(house_prices_regression)
```

```{r}

centers <- as.data.frame(resultado_kmeans$centers)
centers$cluster <- rownames(resultado_kmeans$centers)

ggparcoord(centers, columns = 1:10 , groupColumn = 11) 
```






















```{r}

zscore_price <- (house_prices_regression$target_price_eur-mean(house_prices_regression$target_price_eur))/sd(house_prices_regression$target_price_eur)
zscore_area <- (house_prices_regression$SQUARE_MT-mean(house_prices_regression$SQUARE_MT))/sd(house_prices_regression$SQUARE_MT)

zscore_house_prices <- house_prices_regression
zscore_house_prices$target_price_eur <- zscore_price 
zscore_house_prices$SQUARE_MT <- zscore_area

resultado_kmeans2 <- kmeans(zscore_house_prices, centers = 3)
zscore_house_prices$cluster <- as.character(resultado_kmeans2$cluster)

ggplot(zscore_house_prices) +
    aes(y = zscore_house_prices$target_price_eur, x = zscore_house_prices$SQUARE_MT  , colour = cluster) +
    geom_point(shape = "circle", size = 1.5) +
    scale_color_hue(direction = 1) +
    theme_minimal() +
    coord_cartesian(xlim = c(0, 0.5),ylim = c(0, 20))


```



```{r}
resultado_kmeans <- kmeans(scaled_house_prices, centers = 3)
scaled_house_prices$cluster <- as.character(resultado_kmeans$cluster)


ggplot(scaled_house_prices) +
    aes(y = scaled_house_prices$target_price_eur, x = scaled_house_prices$SQUARE_MT  , colour = cluster) +
    geom_point(shape = "circle", size = 1.5) +
    scale_color_hue(direction = 1) +
    theme_minimal() +
    coord_cartesian(ylim = c(0, 0.4),xlim = c(0.00025, 0.001))
```
































```{r}
#BHK.NO

table(house_prices$BHK_NO.)

#Visualization

ggplot(house_prices) +
 aes(x = "", y = BHK_NO.) +
 geom_boxplot(fill = "#112446") +
 labs(x = "Box plot", 
 y = "Number of rooms") +
 theme_bw()

ggplot(house_prices) +
 aes(x = BHK_NO.) +
 geom_density(adjust = 6L, fill = "#112446") +
 labs(x = "Box plot", 
 y = "Number of rooms") +
 theme_bw()

ggplot(house_prices) +
 aes(x = BHK_NO., y = target_price_eur) +
 geom_point(shape = "circle", size = 1.5, 
 colour = "#112446") +
 labs(x = "Number of rooms", y = "Price (Eur)") +
 theme_bw()
#######################################################################################


#shapiro test

hp_sample <- house_prices %>% sample_n(4999)   

scaled_house_prices <- house_prices %>% 
  mutate(BHK_NO. = (BHK_NO. - min(BHK_NO.))/(max(BHK_NO.) - min(BHK_NO.)))


df <- data.frame(table(scaled_house_prices$BHK_NO.))
df
View(scaled_house_prices)

```

```{r}
#SQUARE_MT

hp_sample_featured <- hp_sample %>%
  filter(SQUARE_MT >= 1L & SQUARE_MT <= 500 | is.na(SQUARE_MT))

da <- hp_sample_featured %>%
  group_by(SQUARE_MT) %>%
  arrange(hp_sample_featured) 
 

ggplot(hp_sample_featured) +
  
  aes(x = SQUARE_MT) +
  geom_histogram(bins = 37L, fill = "#112446") +
  labs(
    title = "Distrubiton of Square_MT in random sample (N=4999)"
  ) +
  theme_bw()




min(house_prices$SQUARE_MT)
max(house_prices$SQUARE_MT)

scaled_house_prices3 <- scaled_house_prices2 %>% 
  mutate(SQUARE_MT = (SQUARE_MT - min(SQUARE_MT))/(max(SQUARE_MT) - min(SQUARE_MT)))

View(scaled_house_prices3)

#ADDRESS
################################################################
mean(scaled_house_prices2$ADDRESS)
sd(hscaled_house_prices2$ADDRESS)
zscore_address <- (scaled_house_prices2$ADDRESS - mean(scaled_house_prices2$ADDRESS))/sd(scaled_house_prices2$ADDRESS)
zscore_address %>%
  .[1:100]

mean(zscore_address)
test=sample(scaled_house_prices3$ADDRESS,4999)
View(scaled_house_prices3)
#################################################################
View(scaled_house_prices3)
View(scaled_house_prices2)
View(scaled_house_prices1)
View(house_prices)

scaled_house_prices3 <- house_prices %>%
  mutate(POSTED_BY  = if_else(POSTED_BY  == "1", 622/29451 , if_else(POSTED_BY == "2", 10538/29451, 18291/29451)))
View(scaled_house_prices3)

summary(scaled_house_prices3)

scaled_house_prices3 <- house_prices %>% 
  mutate(BHK_NO. = (BHK_NO. - min(BHK_NO.))/(max(BHK_NO.) - min(BHK_NO.)))

scaled_house_prices3$ADDRESS<- (scaled_house_prices3$ADDRESS-min(scaled_house_prices3$ADDRESS))/(max(scaled_house_prices3$ADDRESS)-min(scaled_house_prices3$ADDRESS))



```





















#VER O QUE É CROSS VALIDATION??
```{r}
#ADDRESS
starwars_cut <-
  starwars %>%
  select(name,height)%>%
  .[1,10]%>%
  mutate(kfold=sample(rep(1:3,10),nrow(.)))

mse<-NULL
for(i in 1:3)
{
  teste<- starwars_cut%>%filter(kfold==i)
  treino <- starwars_cut%%filter(kfold!=i)
  
  mes[i]<-runif(1)
}
```


#VISUALIZATION
```{r}
str(house_prices$target_price_eur)

house_prices + ggplot(geom_violin(aes(group=target_price_eur)))

boxplot(house_prices$target_price_eur, main = "Prices Distribution Plot", ylab = "Values")
```
```{r}
# Install and load the vioplot package
install.packages("vioplot")
library(vioplot)
```

# Assuming you have a dataset named 'df' and want to plot the quartile distribution of the column 'column_name'
```{r}

vioplot(house_prices$target_price_eur, col = "skyblue", main = "Prices Distribution Plot", ylab = "Values")

# Assuming you have a dataset named 'house_prices' and want to create a violin plot for the 'target_price_eur' column
vioplot(house_prices$target_price_eur, col = "skyblue", main = "Prices Distribution Plot", ylab = "Values")

# Remove scientific notation from the y-axis
options(scipen = 999)

# Plot the violin plot again
vioplot(house_prices$target_price_eur, col = "skyblue", main = "Prices Distribution Plot", ylab = "Values", yaxt = "n")



# Assuming you have a dataset named 'house_prices' and want to create a bar plot for the 'target_price_eur' column
barplot(house_prices$target_price_eur, col = "skyblue", main = "Prices Distribution Plot", ylab = "Values")

# Remove scientific notation from the y-axis
options(scipen = 999)

# Get the y-axis labels without scientific notation
y_labels <- formatC(house_prices$target_price_eur, format = "f", digits = 0)

# Plot the bar plot again with adjusted y-axis labels
barplot(house_prices$target_price_eur, col = "skyblue", main = "Prices Distribution Plot", ylab = "Values", names.arg = y_labels)














str(house_prices)

library(ggplot2)




# Assuming you have a dataset named 'house_prices' and want to create a violin plot for the 'target_price_eur' column
ggplot(house_prices, aes(x = "", y = target_price_eur)) +
  geom_violin(fill = "skyblue", color = "black") +
  labs(title = "Prices Distribution Plot", y = "Values") +
  theme_minimal()


numeric_columns <- house_prices[, sapply(house_prices, is.numeric)]
correlation <- cor(numeric_columns, use = "pairwise.complete.obs")

# Create the heatmap
heatmap(correlation, main = "Correlation Heatmap")

# Create the correlation plot with diagonal values only
corrplot(correlation, type = "lower", order = "hclust")


shapiro.test(sample(house_prices$target_price_eur,4999))


#view(dataset)
library(tidyverse)
#na
library(GGally)
library(dbscan)
# resume select columns
#filter
#
#str_detect
# res_aov , curva normaç
# curva da Anova

# correlação revenue e score
#profit em percentagem

# normalizar..

# sales tipo 3


# Create the table of frequencies
table_posted_by <- table(house_prices$POSTED_BY)

# Create the barplot
barplot(table_posted_by)

# Calculate the percentages
percentages <- round((table_posted_by / sum(table_posted_by)) * 100, 1)

# Add percentage labels to the bars
text(x = barplot(table_posted_by),main = "Category marking who has listed the property", y = table_posted_by, labels = paste0(percentages, "%"), pos = 3)



# cluster segundo dois métodos...
# análisar cluster, relacionar com as features..

#usaram revenue and the budget, relações entre observações e não variáveis...Encontrar perfis tipo coisas em comum
# profits


```


